Learning Network Size while Training with ShrinkNets

### 2018-02-08 paper/ICML/approach.tex
First pass on section 2

\begin{itemize}
  \item Assign an on/off switch to each neuron. But this is np-hard, so we
  consider a relaxation
  \item Formulation
  \item Relationship to group sparsity
  \item Strategies to kill neurons (relation to theory above?)
  \item Implementation details
\end{itemize}

Our approach has two main challenges. First, we need a way to
dynamically size the network during training. Second, we need a
loss function that optimizes for the additional task of sizing, without
deteriorating the learning performance of the main task. Our
approach, called ShrinkNets, copes with both challenges.

During training, our approach starts
with an explicitly over-sized network. As training progresses, we learn
which neurons are not contributing to learning and remove them
dynamically, effectively shrinking the network. This method requires two key components: first, we need a way to identify neurons that are not contributing to the
learning process, and second we need a way to balance the network size and the
generalization capability for the main task. We introduce a new \textsf{Filter} 
\mpv{Choose a different name if possible bc this can be confused w/convolutional
filters. Maybe Shrinkage layer?} 
layer that takes care of \emph{deactivating} neurons. We also modify
existing loss functions to incorporate a new term that takes care of balancing
network size and generalization capability appropriately.

\textbf{Filter Layers: } Filter layers have weights in the range $[0,+\infty]$ and are usually placed after
linear and convolutional layers.
The \textit{Filter Layer} takes an input of size $\left(B \times C \times D_1
  \times \dots \times D_n\right)$, where $B$ is the batch size, $C$ the number
of features (or channels, in the case of convolutional layers), and $D$ any
additional dimension. This structure makes it compatible with fully connected
layers with $n=0$ or convolutional layers with $n=2$. Their crucial property is
a parameter $\theta \in \mathbb{R}^C$. The output is defined as follows: \vspace{-1em}
\begin{equation} Filter(I;\theta) = I \circ \max(0, \theta) \end{equation}
Where $\circ$ is the pointwise multiplication, and $\theta$ is expanded in all
dimensions to match the input size (except the second one since they are equal
by definition). It is easy to see that if for any $k$, if $\theta_k \leq 0$,
the $k^{\text{th}}$ input feature/channel is multiplied by zero and have no
influence on the output. If this happens, we say the Filter layer deactivates
the neuron. These disabled neurons/channels can be removed from the network
without changing its output. Before explaining how that is achieved, we explain
next how the weights of the Filter Layer are initialized and adjusted during
training.

\textbf{Training Procedure: } Once Filter layers are placed in a network and initialized (sampled from the Uniform$[0, 1]$ distribution),
%To train networks we need start with a substantially oversized network, then we
%insert \textit{Filter Layers}  (usually after every linear or convolutional
%layer except the last one) and we sample their weight from the
%$\text{Uniform}(0, 1)$ distribution. 
we could train the network directly using our standard loss function, and we could achieve performance equivalent to a normal neural
network. However, our goal is to find the smallest network with reasonable
performance. We achieve that by introducing sparsity in the parameters of the
\textit{Filter Layers}, thus forcing the deactivation of neurons%. Indeed, having a negative component in the $\theta$
%parameter of the filter layer permamently disable its associated feature
%\gl{Maybe redundant ? we talked about that in the previous paragraph} . 
. To obtain this sparsity, we simply redefine the loss function:
\vspace{-.5em}
\begin{equation}
  L'(x,y;\theta) = L(x, y) + \lambda|\max(0, \theta)|
\end{equation}

The additional term $\lambda|\max(0, \theta)|$ introduces sparsity (see Lasso
loss~\cite{Tibshirani1996}). 
% The $\lambda$ parameter, that can take any
% positive value, adjusts how aggressively the network deactivates neurons, with
% larger values indicating more aggressive deactivation.
 The second component of the loss increases the gradient with respect to $\theta$, thus pushing its value towards zero. Neurons with little impact
on the original loss (gradient lower than $\lambda$), will not be able to
compete against this attraction towards zero. Because the entries in $\theta$
with a value of $0$ or less correspond to dead neurons, $\lambda$ effectively
controls the number of neurons/channels in the entire network. We introduced
the $\max(\dots)$ into the loss to make sure that neurons are permamently disabled
when performing gradient descent based optimization. Next, we
explain how to implement ShrinkNets efficiently.

\subsection{Strategies to Remove Neurons}

\textbf{Implementation}
It is possible to reduce the overhead of the training process by removing
neurons as soon as they become deactivated by $\theta$ going to 0.
%If
%disabled neurons are not quickly removed, the overhead might cause the training
%process to be significantly slower than classic neural networks. 
To do this, we implemented a \emph{neural garbage collection} mechanism which
prunes deactivated neurons on-the-fly,  reducing the processing time
and memory overhead. To support this feature, it is crucial to understand the
information flow between neurons and layers in the neural network. We achieve
this by representing such information flow as a graph. Vertices represent layers,
and edges are event-hubs responsible for propagating information about disabled
neurons to the relevant layers.

>>>>

from the universality theorem, we know that we can find a network that ca
arbitrarly fit a function. Therefore, for a given error tolreance there exists
an optimal size to solve a given problem.

If we have an oversized network then there exist a pruned version that can still
achieve our goal. The main idea is to consider an on/off switch for each neuron,
and we want to find an assignement for this switches that achieve a certain
size/accuracy tradeof. We model these on/off switches by multiplying each input
(or output) of each layer by a number $0$ or $1$. $0$ will deactivate the
neuron, $1$ will let the signal go through. We want to minimize the number of on
switches to reduce the model size as much as we can. This can be modeled solved
by jointly minimizing the objective of the network and a factor of the L0 norm
of the vector containing all the on/off switches.

Finding an optimal binary assignement is an NP-Hard problem (\textbf{Should we
prove this ? I think it should be fairly doable reducing it to 3-SAT
considering the structure of neural networks}). We decide to relax this
problem because this is what people usually do with NP-Hard problem. Our
relaxation is we allow $\bm{\theta}$ to be a real number instead of a boolean We
also use L1 instead of L1. This way we obtain a non-convex, but at least
differentiable (almost everywhere).

This approach assumes that we start with an upper bound on the model size. This
obviously translates in a computational overhead. Our insight is that some
usless neurons (we have multiple definitions below) can be removed early without
impacting the final solution. It has two practical implications: It mitigate the
issue we describe but it also allows other neurons to adapt as soon as one of
their peer is killed. Existing technique usually remove them after convergence
and require an extra fine-tuning process to compensate for the removal.

The key components of the system are: The filter vectors that simulate the
continuous on/off switches, a regularization that tries to kill neurons, the
neuron removal strategy that detects neurons that should probably be removed,
and the garbage collection that effectively remove dead neurons from the model,
and the simplifaction procedure that remove the filter vector for fast
inference.

We will describe these components in the upcoming sections.

\subsection{Notations}

\par In order to avoid any potential ambiguity, in this section we will
describe in details the mathematical notations used in this article. Non-bold
letters represent scalar values, while bold lowercase and upper case
repectively denote vectors and matrices. $\bm{A}^T$ stands for the transpose of
the matrix $\bm{A}$. Subscripts are used to index particular elements of
vectors and matrices. $\bm{x}_i$, $\bm{A}_i$, $\left(\bm{A}^T\right)_j$ and
$\bm{A}_{i,j}$ respectively correspond to the $i^{th}$ component of $\bm{x}$,
the $i^{th}$ row of $\bm{A}$, the $j^{th}$ column of $\bm{A}$ and the $j^{th}$
component of the $i^{th}$ row of $\bm{A}$. All the following definitions assume
$\bm{A}$ to be an $n\times p$ matrix.  For any vetor $\bm{b}$ with $n$
components, we define $\diag{\bm{b}}$ a $n\times n$ matrix such that: $\forall
1 \leq i \leq n$, $\diag{\bm{b}}_{i, i} = \bm{b}_i$ and $0$ otherwise.  For any $l \in \left[0, +\infty\right]$ we define the norm: $\norm{\bm{A}}_l =
\left(\sum_{i=1}^n \sum_{j=1}^p \abs{\bm{A}_{i, j}}^l\right)^{\frac{1}{p}}$. For the rest of this paper and unless stated otherwise, $\bm{y}$ will represent the output of a network, $\bm{x}$ the input, $\bm{b}$ a bias, $\lambda$ regularization factors, $\Omega$ regularization methods, $\bm{\theta}$ general model parameters and $a$ will stand for any element-wise activation function. The only constraint that we want to enforce is that $a(0) = 0$. We use $\intint{u, v}$ to denote inteveral of integers, $\bm{0}$ is the null vector (size depending on the context). $\#S$ is meant to represent the cardinality of a set $S$. To simplify the notation of function composition we use the following operator: $g(f(\bm{x})) = (f \circ g)(\bm{x})$ and for a long sequence of functions from $f_1$ to $f_n$ we use: $f_n(...f_1(\bm{x})) = \left(\bigcirc_{k = 1}^n f_k\right)(\bm{x})$.

\subsection{The Switch Layer}

\textbf{Do you like the name of the layer ?}

Switch layers have weights in the range $[-\infty,+\infty]$ and are usually placed after
linear and convolutional layers.
The \textit{Switch Layer} takes an input of size $\left(B \times C \times D_1
  \times \dots \times D_n\right)$, where $B$ is the batch size, $C$ the number
of features (or channels, in the case of convolutional layers), and $D$ any
additional dimension. This structure makes it compatible with fully connected
layers with $n=0$ or convolutional layers with $n=2$. Their crucial property is
a parameter $\theta \in \mathbb{R}^C$. The output is defined as follows: \vspace{-1em}
\begin{equation} Switch(\bm{I};\bm{\theta}) = \diag{\bm{\theta}} \bm{I}  \end{equation}
Where $\theta$ is expanded in all
dimensions to match the input size (except the second one since they are equal
by definition). It is easy to see that if for any $k$, if $\theta_k \leq 0$,
the $k^{\text{th}}$ input feature/channel is multiplied by zero and have no
influence on the output. If this happens, we say the Switch layer deactivates
the neuron. These disabled neurons/channels can be removed from the network
without changing its output. Before explaining how that is achieved, we explain
next how the weights of the Switch Layer are initialized and adjusted during
training.

\subsection{Training Procedure} Once Switch layers are placed in a network and
initialized (sampled from the $\mathcal{N}(0, 1)$ distribution),
%To train networks we need start with a substantially oversized network, then we
%insert \textit{Switch Layers}  (usually after every linear or convolutional
%layer except the last one) and we sample their weight from the
%$\text{Uniform}(0, 1)$ distribution. 
we could train the network directly using our standard loss function, and we could achieve performance equivalent to a normal neural
network. However, our goal is to find the smallest network with reasonable
performance. We achieve that by introducing sparsity in the parameters of the
\textit{Switch Layers}, thus forcing the deactivation of neurons%. Indeed, having a negative component in the $\theta$
%parameter of the filter layer permamently disable its associated feature
%\gl{Maybe redundant ? we talked about that in the previous paragraph} . 
. To obtain this sparsity, we simply redefine the loss function:
\begin{equation}
  L'(\bm{x},\bm{y};\bm{W}, \bm{\theta}) = L(\bm{x}, \bm{y}; \bm{W}) +
  \lambda\norm{\theta}_1 + \lambda_2\norm{\bm{W}}_p
\end{equation}

The additional term $\lambda|\max(0, \theta)|$ introduces sparsity (see Lasso
loss~\cite{Tibshirani1996}). 
% The $\lambda$ parameter, that can take any
% positive value, adjusts how aggressively the network deactivates neurons, with
% larger values indicating more aggressive deactivation.
 The second component of the loss increases the gradient with respect to $\theta$, thus pushing its value towards zero. Neurons with little impact
on the original loss (gradient lower than $\lambda$), will not be able to
compete against this attraction towards zero. Because the entries in $\theta$
with a value of $0$ or less correspond to dead neurons, $\lambda$ effectively
controls the number of neurons/channels in the entire network. Without the last
term our problem sounds very similar to the Group Sparsity regularization which
is well known in the area of linear and logistic regressions. In the next
section we will try to undercover the relationship between these two problems,
explain why we need this additional regularization term and what should be the
value of $p$.

\input{theory}

\subsection{Speeding up training}

In order to have a practical implementation of ShrinkNets it is absolutely
necessary to prune useless neurons as quick as possible. In order to try more
different strategies we split this task in two subproblems: detecting
potentially non-crucial neurons and setting their corresponding entry in the
switch layer to exactly 0, and neural garbage collection that

\subsubsection{Neuron killing}
\textbf{Threshold strategy} \\ We kill neurons based on a threshold (in absolute
value), this is the method used by deep compression. It is bad because it is
scale dependant and is not robust to noise in the gradients (explain that)
\textbf{Sign change strategy} \\ We look at when the sign change. This is also
sensitive to noise but has the advantage
\textbf{Sign variance strategy}

In our library we implemented a variant of the second strategy and the last one.

\subsubsection{Neural garbage collection}

It is possible to reduce the overhead of the training process by removing
neurons as soon as they become deactivated by $\theta$ going to 0.
%If
%disabled neurons are not quickly removed, the overhead might cause the training
%process to be significantly slower than classic neural networks. 
To do this, we implemented a \emph{neural garbage collection} mechanism which
prunes deactivated neurons on-the-fly,  reducing the processing time
and memory overhead. To support this feature, it is crucial to understand the
information flow between neurons and layers in the neural network. We achieve
this by representing such information flow as a graph. Vertices represent layers,
and edges are event-hubs responsible for propagating information about disabled
neurons to the relevant layers. As soon as a layer receive an event, it updates
his parameters to reflect the new size. Each operation is stored in a log that
is used later to update other component of the system. For example some
optimizers like Adam [Ref] store state about each parameter, their state needs
to be updated the exact same way and at the exact same time as layers in order
to obtain apropriate parameter updates.

\subsection{Speeding up Inference}

The sole puropose of \textit{Switch Layer} are to detect which neurons should be
killed and which should be left. When the training is over we will never remove
more neurons. Therefore, they loose all their interest at inference time. To
improve size and inference time we propose the following method to get rid of
them without changing the output of the model (modulo floating point errors).
The basic idea is to start from every \textit{Switch Layer} and try to find the
nearset linear operator (child or parent) and multiply its weights by the values
in the switch layer. For example we can merge Filter layers with surrounding
Convolutional layer, Linear layer or even BatchNormalization [ref] if they are
before the \textit{Switch Layer}. However we need to be extremely careful not to
cross a non-linearity otherwise it would change the output of network.
